---
title: "Data cleaning/processing"
output: 
  html_document:
    toc: true
    toc_float: true
--- 

```{r,include=FALSE,message=FALSE,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
library(plotly)
library(highcharter)
library(rworldmap)
library(gganimate)
library(transformr)
library(corrplot)
library(leaps)
library(kableExtra)
library(modelr)
library(mgcv)
library(magick)
library(readr)

knitr::opts_chunk$set(
echo = FALSE,
fig.width = 7,
fig.height = 5,
fig.asp = 0.6,
out.width = "60%")
theme_set(theme_bw() +
         theme(legend.position = "bottom",
               legend.title = element_blank(),
               plot.title = element_text(hjust = 0.5, size = 15),
               plot.subtitle = element_text(hjust = 0.5, size = 12)))
```

### Data Import and Cleaning



#### HIV Deaths By Country Dataset

```{r, message = FALSE,warning = FALSE}
deaths = read_csv("data/final_hiv_deaths.csv") %>%
 janitor::clean_names() %>%
 select("location_id", "location_name", starts_with("sex"), starts_with("age"), "year", "val", "upper", "lower") %>%
 rename(country_name = location_name) %>% 
  mutate(year = as.character(year))

deaths %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(deaths),
           n_missing = map(deaths, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```

The death dataset contains HIV death count data from 204 distinct countries, in the 30 years between 1990 and 2019.

The deaths dataset contains `r nrow(deaths)` rows and `r ncol(deaths)` columns, with each row representing a single record of number of HIV deaths in a country stratified by year, sex and age group. Variables in the deaths dataset are as follows: `r colnames(deaths)`. 


#### GDP Per Capita By Country Dataset

```{r, message = FALSE,warning = FALSE}
gdp = read_csv("data/final_gdp_per_capita.csv", skip = 4, col_names = TRUE) %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 

names(gdp) <- sub('^x', '', names(gdp))

gdp_clean = gdp %>%  
  select("location_id", "country_name", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "GDP_per_capita") 

gdp_clean %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(gdp_clean),
           n_missing = map(gdp_clean, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```
The raw GDP dataset contains GDP per capita (US dollars) from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(gdp)` rows and `r ncol(gdp)` columns, with each row representing a single record of a country's GDP per capital (US dollars) in a given year. 


#### Population By Country Dataset
```{r, message = FALSE,warning = FALSE}
population = read_csv("data/final_population_total.csv", skip = 4, col_names = TRUE)  %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 

names(population) <- sub('^x', '', names(population))

population_clean = population %>%  
  select("location_id", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "Population") 

population_clean %>%
 summarize(n_countries = n_distinct(location_id),
           n_years = n_distinct(year),
           col_name = colnames(population_clean),
           n_missing = map(population_clean, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```

The raw population dataset contains total population from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(population)` rows and `r ncol(population)` columns, with each row representing a single record of a country's population in a given year. 



### Merging 3 datasets to produce the "maindata" dataset for visualization and analysis

First, we merge the GDP Per Capita By Country dataset (**gdp_clean**) and Population By Country Dataset (**population_clean**) into a dataset called **gdp_population**. This combined dataset contains data from 200 distinct countries between 1990 and 2019.

```{r, message = FALSE,warning = FALSE}
gdp_population = inner_join(gdp_clean, population_clean, by = c("location_id", "year")) %>%
 janitor::clean_names() 

gdp_population %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(gdp_population),
           n_missing = map(gdp_population, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```

Next, we merge the **gdp_population** dataset with the HIV Deaths dataset (**deaths**) into a **maindata** dataset by matching on the countries and years and using `left_join`. We save this as a CSV for further analysis.

```{r, message = FALSE,warning = FALSE}
maindata = left_join(gdp_population, deaths, by = c("year","location_id")) %>% 
  select(-country_name.x, -age_id) %>% 
  rename(country_name = country_name.y) %>% 
  select(location_id, country_name, everything())

maindata %>% 
  summarize(n_countries = n_distinct(location_id),
            n_years = n_distinct(year),
            col_name = colnames(maindata),
            n_missing = map(maindata, ~sum(is.na(.)))) %>% 
  unnest(n_missing)

write_csv(maindata, 'data/maindata.csv')
```

**maindata** has `r nrow(maindata)` observations and `r ncol(maindata)` features. Basic features we are interested in include:

- `location_id`: id corresponding to a country_name
- `sex_id`: id corresponding to sex
- `sex_name`: Male or Female
- `age_name`: Aae groups including "0-9 years", "10-24 years",  "25-49 years", "50-74 years", and "75+ years".
- `year`: Year of the records
- `val`: Point estimate of number of HIV/AIDS deaths
- `upper`: Upper estimate of number of HIV/AIDS deaths
- `lower`: Lower estimate of number of HIV/AIDS deaths
- `gdp_per_capita`: GDP per capita in 2022 US dollars
- `population`: population total
- `country_name`: name of country


