---
title: "Data cleaning/processing"
output: 
  html_document:
    toc: true
    toc_float: true
--- 

```{r,include=FALSE,message=FALSE,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
library(plotly)
library(highcharter)
library(rworldmap)
library(gganimate)
library(transformr)
library(corrplot)
library(leaps)
library(kableExtra)
library(modelr)
library(mgcv)
library(magick)

knitr::opts_chunk$set(
echo = FALSE,
fig.width = 7,
fig.height = 5,
fig.asp = 0.6,
out.width = "60%")
theme_set(theme_bw() +
         theme(legend.position = "bottom",
               legend.title = element_blank(),
               plot.title = element_text(hjust = 0.5, size = 15),
               plot.subtitle = element_text(hjust = 0.5, size = 12)))
```

### Data Import and Cleaning



#### Death dataset.

```{r}
deaths = read_csv("data/final_hiv_deaths.csv") %>%
 janitor::clean_names() %>%
 select("location_name", starts_with("sex"), starts_with("age"), "year", "val", "upper", "lower") %>%
 rename(country_name = location_name) %>% 
  mutate(year = as.character(year))

deaths %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(deaths),
           n_missing = map(deaths, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```



The death dataset contains HIV death count data from 204 distinct countries, in the 30 years between 1990 and 2019.

The deaths dataset contains `r nrow(deaths)` rows and `r ncol(deaths)` columns, with each row representing a single record of number of HIV deaths in a country stratified by year, sex and age group. Variables in the deaths dataset are as follows: `r colnames(deaths)`. 

The death dataset contains HIV death count data from 204 distinct countries, in the 30 years between 1990 and 2019.


#### GDP dataset

```{r}
gdp = read_csv("data/final_gdp_per_capita.csv", skip = 4, col_names = TRUE) %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 

names(gdp) <- sub('^x', '', names(gdp))

gdp_clean = gdp %>%  
  select("location_id", "country_name", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "GDP_per_capita") 

```
The raw GDP dataset contains GDP per capita (US dollars) from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(gdp)` rows and `r ncol(gdp)` columns, with each row representing a single record of a country's GDP per capital (US dollars) in a given year. Variables in the GDP dataset are as follows: `r colnames(gdp)`. 

### Population dataset
```{r}
population = read_csv("data/final_population_total.csv", skip = 4, col_names = TRUE)  %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 

names(population) <- sub('^x', '', names(population))

population_clean = population %>%  
  select("location_id", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "Population") 
```

The raw population dataset contains total population from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(population)` rows and `r ncol(gdp)` columns, with each row representing a single record of a country's population in a given year. Variables in the deaths dataset are as follows: `r colnames(gdp)`. 

```{r}
gdp_population = inner_join(gdp_clean, population_clean, by = c("location_id", "year")) %>%
 janitor::clean_names() 
```





