---
title: "Regression Analysis"
always_allow_html: true
output: 
  html_document:
    toc: true
    code_folding: hide
    toc_float: true
--- 

```{r,include=FALSE,message=FALSE,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
library(plotly)
library(highcharter)
library(rworldmap)
library(gganimate)
library(transformr)
library(corrplot)
library(leaps)
library(kableExtra)
library(modelr)
library(mgcv)
library(magick)
library(readr)
library(gifski)

knitr::opts_chunk$set(
echo = FALSE,
fig.width = 7,
fig.height = 5,
fig.asp = 0.6,
out.width = "60%")
theme_set(theme_bw() +
         theme(legend.position = "bottom",
               legend.title = element_blank(),
               plot.title = element_text(hjust = 0.5, size = 15),
               plot.subtitle = element_text(hjust = 0.5, size = 12)))

maindata = read_csv("data/maindata.csv") 
```


#### Create "mortality_per_100k" and "log_mortality_per_100k" variables

**maindata** has `r nrow(maindata)` observations and `r ncol(maindata)` features. We are adjusting for population size by dividing  `val` (point estimate of number of HIV/AIDS deaths) by `population` multiplied by 100,000 to get a `mortality_per_100k` variable and a `log_mortality_per_100k` variable.

```{r, echo=TRUE, message = FALSE,warning = FALSE}
regression = maindata %>% 
  mutate(mortality_per_100k = val / (population/100000),
         log_mortality_per_100k = log(mortality_per_100k),
         year_2004 = if_else(year <= 2004, 1, 0),
         year_2004 = as.factor(year_2004),
         country_name = as.factor(country_name)) 
regression %>% 
  summarize(col_name = colnames(regression),
            n_missing = map(regression, ~sum(is.na(.)))) %>% 
  unnest(n_missing)
```

### Plots of the distribution to main outcome of interest(HIV mortality per 100k population).

First, we plot the distribution of `mortality_per_100k`. We can see that we need to transform it to satisfy the normality assumption. So we used a log transformation to get a more normal distribution, and have a new variable `mortality_per_100k`.

```{r, echo=TRUE, message = FALSE,warning = FALSE}
regression %>%
 ggplot(aes(x = mortality_per_100k)) +
 geom_histogram(color = "darkblue", fill = "lightblue") +
 ggtitle("Histogram of HIV deaths per 100k") +
 geom_vline( aes(xintercept = mean(mortality_per_100k)),
            linetype = "dashed") +
 ylab("") +
 xlab("Estimated HIV deaths per 100k")

regression %>%
 ggplot(aes(x = log_mortality_per_100k)) +
 geom_histogram(color = "darkblue", fill = "lightblue") +
 ggtitle("Histogram of log(HIV deaths per 100k)") +
 geom_vline(aes(xintercept = mean(log_mortality_per_100k)),
            linetype = "dashed") +
 ylab("") +
 xlab("Estimated log(HIV deaths per 100k)")
```

### Linear regression models

### Hypothesis 1:
There is statistically significant relationship between the HIV deaths per 100k and age groups, sex, year, and the country's gdp respectively. 

### model 1:

$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4 gdp $$
```{r}
regression_logit = regression %>%
  mutate(log_mortality_per_100k = if_else(log_mortality_per_100k == -Inf, 0.01, log_mortality_per_100k))
  
model1_logit = lm(log_mortality_per_100k ~ age_name + sex_name + year + gdp_per_capita, data = regression_logit) %>% 
  broom::tidy() %>% 
  mutate(exp_est = exp(estimate))

knitr::kable(model1_logit)

```

Interpretation:
The 10-24 years old has 0.83 times the HIV mortality per 100k population compared to the 0-9 years old holding sex, year and gdp in constant.
The 25-49 years old has 9.26 times the HIV mortality per 100k population compared to the 0-9 age years old holding sex, year and gdp in constant.
The 50-74 years old has 2.70 times the HIV mortality per 100k population compared to the 0-9 age years old holding sex, year and gdp in constant.
The 75+ years old has 0.06 times the HIV mortality per 100k population compared to the 0-9 age years old holding sex, year and gdp in constant.
The HIV mortality per 100k population will increase by 4% with 1 year increase between 1990-2019 holding age, sex, and gdp in constant.
The the HIV mortality per 100k population will decrease by 0.004% with 1 unit increase of gdp holding age, sex, and year in constant.

The associations between the HIV deaths per 100k and age groups, sex, year, and the country's gdp respectively are all statistically significant at the level of 5%.


### Hypothesis 2:
There is a significant change of the HIV deaths per 100k before and after the year 2004 controlling for age groups, sex, and gdp.

### model 2:
$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year_{2004} + \beta_4gdp  $$
```{r}
  
model2_logit = lm(log_mortality_per_100k ~ age_name + sex_name + year_2004 + gdp_per_capita, data = regression_logit) %>% 
  broom::tidy() %>% 
  mutate(exp_est = exp(estimate))

knitr::kable(model2_logit)

```
Interpretation: 
The HIV deaths per 100k is 44.51% lower before 2004 than after 2004 after controlling for age, sex, and gdp at the level of 5% significance. 

### Hypothesis 3:
There is an interaction between age and sex. 

### model 3:

$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4 gdp + \beta_5 age*sex $$

```{r}
model1_compare = lm(log_mortality_per_100k ~ age_name + sex_name + year + gdp_per_capita, data = regression_logit)

model3_logit = lm(log_mortality_per_100k ~ age_name + sex_name + year + gdp_per_capita + age_name*sex_name, data = regression_logit) 

anova(model1_compare, model3_logit) 

```

Interpretation:
There is an interaction between age groups and sex at the 5% significance level, which means that gender can have different impacts on the HIV deaths per 100k regarding to different age groups.
